---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

Transcription factor Occupancy Profiler (TOP)
=================================================

<!-- badges: start -->
<!-- badges: end -->

Transcription factor Occupancy Profiler (TOP) fits a Bayesian hierarchical model 
using transcription factor (TF) motifs, DNase- or ATAC-seq data,
as well as ChIP-seq data (only required in training) 
from multiple TFs across multiple cell types. 

It can be used to predict the quantitative occupancy or binding probability
for many TFs using data from a single DNase- or ATAC-seq experiment. 
Thus, it allows efficient profiling of quantitative TF occupancy landscapes 
across multiple cell types or conditions using DNase- or ATAC-seq experiments.

[TOP R package website](https://kevinlkx.github.io/TOP/) 
------------------------

Install R package
```{r install-R-package, eval=FALSE}
install.packages("devtools")
library(devtools)
devtools::install_github("kevinlkx/TOP")

library(TOP)
```

Reference
---------

* Luo, K., Zhong, J., Safi, A., Hong, L., Tewari, A., Song, L., Reddy, T., Ma, L., 
Crawford, G., & Hartemink, A. (2020) "Quantitative occupancy of myriad 
transcription factors from one DNase experiment enables 
efficient comparisons across conditions." bioRxiv, bioRxiv:2020.06.28.171587.

