---
title: 'Predict TF occupancy using trained TOP model'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predict TF occupancy using trained TOP model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

Load TOP R package
---------------------
```{r load-TOP-package, eval=TRUE, message=FALSE, warning=FALSE}
library(TOP)
```

Load data
-----------

To make predictions using trained TOP models, 
we first need to prepare the input data as a data frame, with six columns.

Columns from left to right are PWM scores and five DNase (or ATAC) bins. 

You can follow this [page](data_preparation.html) to prepare the input data. 

Let's load an example ATAC-seq data to predict CTCF occupancy in K562 cell type. 

This data contains binned ATAC-seq counts around CTCF motif matches in chr1. 
```{r load-combined-data, eval=TRUE, include=FALSE}
data <- readRDS(system.file("extdata/example_data", "CTCF.K562.ATAC.chip.example.data.rds", package = "TOP"))
cols <- c('chr','start','end','name','pwm.score','strand','p.value', paste0('bin', 1:5))
data <- data[cols]
```

```{r, eval=TRUE}
head(data,3)
```

Predict TF occupancy using pretrained TOP model
------------------------------------------------

We can predict TF occupancy for a TF in a cell type using 
the `predict_TOP()` function.

We have trained TOP models for ATAC-seq and DNase-seq (Duke and UW) data using ENCODE data,
and included pre-trained TOP model coefficients (posterior mean) within the package.

If you want to use the pre-trained TOP model coefficients included in the package,
you simply specify the model to use, by setting `use_model` equals "ATAC", 
"DukeDNase", or "UwDNase". 

For example, here we chose the pretrained "ATAC" model.

We chose the "bottom" level regression coefficients 
as we know that have CTCF ChIP data from K562 cell type in our training set. 
```{r predict-TOP-use-pretrained-coef, eval=TRUE}
TOP_result <- predict_TOP(data, 
                          tf_name = 'CTCF', 
                          cell_type = 'K562', 
                          use_model = 'ATAC', 
                          level = 'bottom', 
                          logistic_model = FALSE, 
                          transform = 'asinh')
```

Or, we can let the software to choose the "best" available level of coefficients 
for the TF and cell type of interest. 
In this case, it automatically selected the "bottom" level as the best level.
```{r predict-TOP-mean-coef-best, eval=TRUE}
TOP_result <- predict_TOP(data, 
                          tf_name = 'CTCF', 
                          cell_type = 'K562', 
                          use_model = 'ATAC', 
                          level = 'best', 
                          logistic_model = FALSE, 
                          transform = 'asinh')
```

If you want to use your own model, you can specify that by setting `TOP_coef` to 
your TOP model coefficients. 
Check [this](train_TOP_model.html) and [this](train_TOP_logistic_model.html) 
tutorials to train your own models.

For example, we can load TOP quantitative occupancy model for ATAC-seq data 
included in the packge:
```{r load-TOP-posterior-mean-coef, eval=TRUE}
TOP_coef <- readRDS(system.file("extdata/trained_model_coef/ATAC", "TOP_M5_posterior_mean_coef.rds", package = "TOP"))
```

We can use the pretrained model included in the package
```{r predict-TOP-mean-coef-best-2, eval=TRUE}
TOP_result <- predict_TOP(data, 
                          TOP_coef = TOP_coef, 
                          tf_name = 'CTCF', 
                          cell_type = 'K562', 
                          level = 'best', 
                          logistic_model = FALSE, 
                          transform = 'asinh')
```

This result contains a list with the level of model, 
regression coefficients (posterior mean) used to make predictions, 
and predictions of TF occupancy. 

The prediction result contains the training data together with predicted 
TF occupancy for each candidate site:
```{r}
TOP_predictions <- TOP_result$predictions

head(TOP_predictions, 5)
```

We can plot the predicted occupancy vs. measured occupancy
```{r plot-predicted-measured, eval=TRUE, fig.width=5, fig.height=5}
data_chip <- readRDS(system.file("extdata/example_data", "CTCF.K562.ATAC.chip.example.data.rds", package = "TOP"))

scatterplot_predictions(x = asinh(data_chip$chip),
                        y = asinh(TOP_predictions$predicted),
                        title = 'Predicting CTCF occupancy in K562 cell',
                        xlab = 'asinh(measured occupancy)',
                        ylab = 'asinh(predicted occupancy using bottom level coefficients)',
                        xlim = c(0,8),
                        ylim = c(0,8))
```

We can also try the "middle" level regression coefficients for CTCF, 
which are cell-type generic, thus could be used to predict CTCF in any cell types.
```{r predict-TOP-mean-coef-middle, eval=TRUE}
TOP_middle_result <- predict_TOP(data, 
                                 TOP_coef = TOP_coef, 
                                 tf_name = 'CTCF', 
                                 level = 'middle', 
                                 logistic_model = FALSE, 
                                 transform = 'asinh') 

TOP_middle_predictions <- TOP_middle_result$predictions
```

We can see the performance of the "middle" level model is very close to the "bottom" level.
```{r plot-predicted-measured-middle, eval=TRUE, fig.width=5, fig.height=5}
scatterplot_predictions(x = asinh(data_chip$chip),
                        y = asinh(TOP_middle_predictions$predicted),
                        title = 'Predicting CTCF occupancy in K562 cell',
                        xlab = 'asinh(measured occupancy)',
                        ylab = 'asinh(predicted occupancy using middle level coefficients)',
                        xlim = c(0,8),
                        ylim = c(0,8))
```


Predict TF binding probability using trained TOP logistic model
------------------------------------------------------------------

We can also predict TF binding probability for these candidate sites using 
pre-trained TOP logistic model, by setting `logistic_model = TRUE`.

Predict CTCF binding probability in K562:
```{r predict-TOP-logistic-mean-coef, eval=TRUE}
TOP_result <- predict_TOP(data, 
                          tf_name = 'CTCF',
                          cell_type = 'K562',
                          use_model = 'ATAC',
                          level = 'best',
                          logistic_model = TRUE)
logistic_predicted <- TOP_result$predictions

head(logistic_predicted, 5)
```

